---
title: "Dummy homework"
author: "Name"
output:
  bookdown::pdf_document2:
    number_sections: no
    toc: no
---
Here is a simple \textsf{R} Markdown template for presenting homework exercises. Notice that use of \textsf{R} Markdown is not compulsory.

- Template includes some basic examples about code blocks, equations, figures and cross-referencing.

- Rmd file can be compiled with the key combination \texttt{Ctrl + Shift + K} or by pressing the \texttt{Knit} button on \textsf{R} Studio interface.

- Code chunk can be inserted with the key combination \texttt{Ctrl + Alt + I} or by pressing the \texttt{Insert} button on \textsf{R} Studio interface.

- Equations work very similarly to \LaTeX. However, cross-referencing differs significantly from \LaTeX\ syntax.

- Notice that cross-referencing is not provided in base R Markdown. [\textcolor{blue}{Bookdown}](https://github.com/rstudio/bookdown) package provides output formats that enable cross-referencing.

- For more information about \textsf{R} Markdown, see

  - [\textcolor{blue}{R Markdown: The Definitive Guide}](https://bookdown.org/yihui/rmarkdown/).
  
  - [\textcolor{blue}{R Markdown Cookbook}](https://bookdown.org/yihui/rmarkdown-cookbook/).
  
  - Especially, [\textcolor{blue}{Chapter 1}](https://bookdown.org/yihui/rmarkdown/installation.html) of the book "R Markdown: The Definitive Guide" includes instructions for installation of \textsf{R} Markdown.

# Dummy exercise

## Code blocks

Below is an example code block.
```{r}
emis <- read.table("emissions.txt", header = TRUE, sep = "\t", row.names = 1)
fit <- lm(NOx ~ ., data = emis)
head(emis)
```

## Equations and inline R code 
Value of the *coefficient of determination* is approximately $R^2 \approx$ `r round(summary(fit)$r.squared, 2)`. Previous value was actually computed with inline R command (see Rmd file)! Coefficient of determination can be calculated with the formula

\begin{equation*}
  R^2 = 1 - \frac{SSE}{SST}.
\end{equation*}
Additionally, one can have numbered equations and cross-reference them,
\begin{equation} (\#eq:rsquared)
  R^2 = 1 - \frac{SSE}{SST}.
\end{equation}
Coefficient of determination $R^2$ is calculated according to Equation \@ref(eq:rsquared).

## Plotting

Notice that labeled text is quite useful for making clean figure captions (see Rmd file).

(ref:captionscatter) Scatter plot of variables.

```{r scatter, fig.cap="(ref:captionscatter)"}
pairs(emis, pch = 19, col = "midnightblue", gap = 0, upper.panel = NULL,
      cex.labels = 1)
```

Figure \@ref(fig:scatter) suggests that there is negative correlation between variables \texttt{NOx} and \texttt{Humidity}.

## Working directory

Notice that while compiling \textsf{R} Markdown document, the working directory is by default the location of Rmd file.
```{r}
getwd()
```

See [\textcolor{blue}{Chapter 16.6}](https://bookdown.org/yihui/rmarkdown-cookbook/working-directory.html) of the book "R Markdown Cookbook" for more information.
